{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-20 thesis/quicksand.tex
tweaks to qs chapter in advance of sending out incmoplete ver.

Note that Iterative Deepening is a wrapper algorithm around stateless MC.
A model checker is still used to test each state space, and other reduction techniques such as DPOR
are still applicable in each.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new preemption points reactively based on any runtime analysis.
Here
will focus on run-time data race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
The next section (\cref{sec:quicksand-soundness})
will prove that in fact,
in addition to statically-known synchronization preemption points,
this suffices to provide at least as strong verification guarantees as any other possible preemption point set.

@@

However, in many tests of even modestly-sized programs,
full verification is not feasible,
and focusing on the maximal state space alone is likely to be fruitless.
%
Hence, Iterative Deepening also allows for prioritizing subset jobs
based on number of preemption points, ETA, and whether data race candidates are included among their preemption points.
It relies on state-space estimation \cite{estimation}
to predict which jobs are likely to complete within a reasonable time,
before actually testing a large fraction of interleavings for each.
The overall goal is to decide automatically when to defer testing a state space,
so an inexpert user can provide only her total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.
Quicksand seeks to maximize completed state spaces,
as each one serves as a guarantee that all possible interleavings therein were tested;
\cref{sec:quicksand-discussion} discusses some limitations of this approach.
The next three subsections will show how to schedule these smaller jobs
based on their preemption points and ETAs.

@@

For kernel-level testing, interrupt-disabling is analogous to locking,
so preemptions must also be introduced
just before a disable-interrupt opcode (on x86, {\tt cli})
and just after interrupts are re-enabled (on x86, {\tt sti}).
During data race analysis, {\tt cli} and {\tt sti} are treated as a single global lock
(note that {\tt cli}'d memory accesses can still race with others that have interrupts on).%
\footnote{Some kernels disable preemption without disabling interrupts,
which can be communicated to the MC using manual annotations,
and must be treated similarly.
This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}
Quicksand is configured to begin with
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
$\{yield,cli\}$,
$\{yield,sti\}$,
and $\{yield,lock,$ $unlock,cli,sti\}$.
As a heuristic, it doesn't test every intermediate subset such as $\{lock,sti\}$,
which would result in $2^p$ jobs right off the bat,
although this could potentially be improved in future work (\cref{sec:quicksand-discussion}).

@@

\begin{figure}[t]
        %\includegraphics[width=0.48\textwidth]{dr-jobs-v2.pdf}
        asdfsaf
        % TODO: put the figure form yr snake fight
	\caption[Quicksand incorporates data races as new preemption points at run-time.]
		{Quicksand incorporates data races as new preemption points at run-time,
		by managing the exploration of multiple state spaces,
		communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
                %When an access pair is reported as a data race candidate,
                When a data race is reported,
                a new preemption point is added for each involved access,
                and new jobs are added for later testing,
                corresponding to different combinations of those with the existing preemption points.}
        \label{fig:new-dr-jobs}
\end{figure}

The new state spaces may expose a failure, in which case Iterative Deepening must stop and report a data-race bug,
or complete successfully, indicating a {\em benign} (i.e., false-positive) data race.
They may also uncover a new data-race candidate entirely in some alternate interleaving,
in which case we may iteratively advance to a superset state space which will preempt at both racing access pairs.
Being constrained by a CPU budget,
Iterative Deepening may time out before completing a data race's associated state space,
in which case the data race remains neither confirmed nor refuted.
%Depending on how much burden the implementation wants to impose on the user,
%it may then report it as a
%report a potential false positive that the user must handle
In such cases, Quicksand elects to impose some burden on the user
by reporting it as a potential false positive
and recommend that she investigate it by hand to judge for herself whether it be a bug.
% TODO: fix this to point to partial verif
% TODO (also) make sure that all other quicksand-discussion references get updated if appropriate
\cref{sec:quicksand-discussion} will discuss future opportunities for improving
debugging output in cases of such {\em partial verification}.
However, experience shows that this interactivity pays off:
in the next chapter's educational user study (\cref{sec:education-eval}),
one student reported during the survey that they used this recommendation,
combined with their own intuition,
to find a bug that Quicksand was not able to find alone (\cref{sec:education-reasons-worthwhile}).

@@

Iterative Deepening also accounts for the inherent inaccuracy of ETA estimates.
Line~1 heuristically scales up the time remaining to avoid suspending jobs too aggressively
in case their ETAs are actually overestimated.
Lines~12-15 account for the
%bizarre
possibility that among two suspended jobs,
%given two jobs,
%%$j_1,j_2$,
\GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$}
but
\GetETA{$j_1$} $>$ \GetETA{$j_2$}.
This may seem surprising,
but can often arise because estimates tend to get more accurate over time,
and $j_1$ perhaps ran much longer, on account of being overall smaller,
before becoming suspended.
In such scenarios,
the algorithm heuristically assumes the smaller job's ETA is more accurate,
in order to avoid repeatedly resuming larger jobs briefly only to find that their ETAs keep getting worse and worse
(in other words, it lets us avoid thrashing in Quicksand). % ¯\_(ツ)_/¯

@@

As long as the allocation heap is properly synchronized,
a Pure Happens-Before analysis should identify a happens-before edge
between line~2's {\tt free} and line~4's {\tt malloc},
and it would report no race.
However, the upcoming evaluation will show that Limited Happens-Before retains some advantages over Pure
(\cref{sec:quicksand-eval-bugs}),
so it is useful to be able to automatically suppress data-race candidates
that are certain to end up being false positives when reordered.
Such collisions could instead be avoided with a hacked allocator which never recycles memory,
but this could unacceptably impact performance in {\tt malloc}-heavy tests.

The ability to disregard reallocation false positives is unique to Iterative Deepening.
When limited to a single test execution, suppressing any data race candidate matching this pattern is unsound.
Consider the more unusual program in Figure~\ref{fig:recycle-bug},
in which the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
Here, reordering lines~6-7 before line~3 will allow {\tt x} and {\tt x2} to race.
Discarding the data race report as a false positive after checking just this one execution
would overlook such a bug,
but Iterative Deepening is guaranteed to explore the alternate interleaving,
in which the true data race will show up without {\tt free()} and {\tt malloc()} interposing,
so it is safe to suppress at first, as I will prove in \cref{sec:quicksand-realloc}.
Moreover,
in the context of Iterative Deepening, being able to discard certain data race candidates
allows Quicksand to skip exploring some entire state spaces,
and hence run fewer Landslides overall;
this is analogous to DPOR's ability to skip equivalent interleavings within a single Landslide instance.
Upcoming in the evaluation, \cref{sec:quicksand-eval-verif}'s \Cref{tab:drstatistics}
will show how many redundant state spaces Quicksand is able to prune with this technique.

% TODO smae as above
\begin{figure}[t]
	\begin{tabular}{rll}
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{publish(x);} & \\
		2 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
		3 & \texttt{\hilight{olivegreen}{free}(x);} \\
		4 & & \texttt{x2 = get\_published\_x();} \\
		5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
		6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
		7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
	\end{tabular}
	% TODO: shortver
	\caption{If a single-pass Limited HB analysis discarded candidates matching the malloc-recycle pattern,
it would miss the bug in this adversarial program.}
	\label{fig:recycle-bug}
\end{figure}

@@

{\bf Assumptions.}
The proofs are built on a DPOR definition which assumes sequentially-consistent memory hardware.
All algorithms involved are assumed to operate on a machine model of a single globally-consistent execution trace,
which fundamentally cannot account for memory reordering nondeterminism.
\cref{sec:quicksand-discussion} discusses this limitation further;
for existing work on combining DPOR with relaxed memory, I refer the reader to \cite{tsopso}.
They also assume the Limited Happens-Before definition
for the data-race analysis.
I leave the case for Pure Happens-Before to future work,
although if I may appeal to intuition,
it requires only to show that for any data-race candidate
Limited Happens-Before reports in a given execution,
that Pure Happens-Before does not,
either it will be a false positive,
or the latter will find it in an alternate execution within the same state space,
or the latter will find a different data race that ultimately leads to a bigger state space in which the first one may be found,
much like a generalization of \cref{sec:quicksand-realloc}.

@@

\begin{itemize}
	\item {\bf Single state space, basic DPOR} ({\tt -C}):
		Runs a single instance of Landslide configured to preempt on all statically-known synchronization PPs.
		Corresponds to dBug's approach \cite{dbug-ssv}.
		Never adds any new preemption points based on data race reports.
	\item {\bf Single state space, ICB} ({\tt -I}):
		Runs a single instance of Landslide with preemption points as above,
		but running Iterative Context Bounding with BPOR (\cref{sec:landslide-icb})
		instead of plain DPOR.
		Corresponds to CHESS's approach \cite{chess}.
		Requires either {\tt -C} or {\tt -M} (see below).
	\item {\bf Single state space, preempt-everywhere} ({\tt -0}):
		Runs a single instance of Landslide as above,
		but preempting on every shared memory access, not just synchronization.
		Corresponds to the approach of SPIN \cite{spin} and Inspect \cite{inspect};
		CHESS supports this mode as well with optional compiler instrumentation.
		Requires {\tt -C}; may optionally be combined with {\tt -I}.
	\item {\bf Maximal state space mode} ({\tt -M}):
		Runs the na\"{i}ve version of Iterative Deepening shown in \Cref{alg:algorithm0},
		i.e.,
		immediately abandons any state space whenever a superset of it exists.
		This results in always testing the maximal state space only, with no inherent parallelism,
		and optimizes for the fastest verification when the user has reason to believe no bugs will exist.
		No prior work implements this approach.
		Note that this mode was implemented after \cite{quicksand}'s publication,
		and I will feature it in the evaluation of transactional memory
		(\cref{sec:tm-eval}) rather than in this chapter.
\end{itemize}

@@

%\begin{figure}[p]
%	\begin{tabular}{c}
%        \includegraphics[width=0.71\textwidth]{../proposal/dowefindbugsfaster-v2.pdf} \\
%		\begin{tabular}{p{\textwidth}}
%                (a) Bugs found by Quicksand versus control experiments, measured in CPU time.
%                %as a function of elapsed CPU time.
%                Overall,
%                a more resource-fair comparison than (b),
%                although Quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\
%		\end{tabular}
%                \\
%                \\
%        \includegraphics[width=0.71\textwidth]{../proposal/dowefindbugsfaster-wallclock-v2.pdf} \\
%                %(b) Bugs found by elapsed wall-clock time.
%		\begin{tabular}{p{\textwidth}}
%                (b) Bugs found by Quicksand versus control experiments, measured in wall-clock time.
%                Quicksand is parallelized tenfold; the vertical line indicates its 1 hour limit. \\
%		\end{tabular}
%        \end{tabular}
%	\caption[Bug-finding performance comparison
%	between Quicksand and prior work.]
%	{Bug-finding performance comparison between
%        by several configurations of Quicksand and the single-state-space approaches.
%	Quicksand finds 125\% as many bugs
%        with data-race PPs
%	at the 10-hour mark, compared to the best prior work approach.}
%        \label{fig:dowefindbugsfaster}
%\end{figure}

@@

Beyond finding new bugs and completing full verifications with data-race PPs,
I evaluated Quicksand's performance for classifying data-race candidates in two ways:
its ability to check nondeterministic data races not reachable under a single-pass analysis (\cref{sec:quicksand-pps})
and its ability to suppress reallocation false positives (\cref{sec:quicksand-id-realloc}).
\Cref{tab:drstatistix} presents the results for this section.

@@

\Cref{fig:dr-falsenegs}
%compares the types of data-race candidates necessary to expose each data-race bug in the test suite.
visualizes the difference between single-pass and model-checking-enabled data race analysis.
The first and third series represent the bugs found using preemption points from single-pass data-race candidates only,
% not entirely true, as portend could be given data-race traces from an MC
%but they don't do it in their paper, so i feel comfortable making this claim
i.e., the state-of-the-art approach used by RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}.
The second and fourth series show all data-race bugs Quicksand found,
which includes the former type as well as new bugs involving nondeterministic races.
QS-Limited-HB found a nice 69 data-race bugs in total, % nice
15 of which %could not be found with single-pass data-race candidates alone.
required nondeterministic data-race preemption points to expose.
QS-Pure-HB is even more dependent thereupon, %on nondeterministic data-race PPs,
%requiring nondeterministic data-race PPs
requiring them in 21 cases among its 44 total data-race bugs.
Moreover, although the frequency
of these nondeterministic races varies across the different test cases
(for example, almost all in {\tt broadcast\_test} were nondeterministic; almost none in {\tt mutex\_test}),
they are still at least present in all tests,
meaning it is not just an issue of writing ``better'' test cases to avoid them.

@@

\subsubsection{Avoiding redundant work}

@@

\subsubsection{Preemption point subsets}

@@

In \cref{sec:quicksand-eval-bugs}, I evaluated the state-of-the-art approach's ability to find data-race-induced failures
by configuring a static predicate to preempt on any non-stack memory access
({\tt -0}, see \cref{sec:quicksand-impl-modes}).
This introduced hundreds of new preemption points on each new test execution,
with a prohibitive performance impact.
While this performance could be improved by
relaxing the preemption strategy,
instead using a static or single-pass analysis to find data-race candidate PPs in advance \cite{portend},
that would sacrifice soundness of the verification guarantee, as I showed in \cref{sec:quicksand-eval-nondets}.
However, Quicksand itself could employ static data-race analysis such as RacerX \cite{racerx},
or single-pass dynamic analysis such as ThreadSanitizer \cite{tsan} in future work.
% These tools tend to err on the side of false positives,
% but also include .....
Any data-race candidates identified in advance could heuristically be included in Quicksand's initial seed PP sets
(\cref{sec:quicksand-initial-pps}),
enabling it to focus on the most suspicious races immediately,
rather than waiting for them to be identified after potentially many iterations of model checking.

\subsubsection{Partial verification}

Quicksand is not the first concurrency tester to provide a partial verification guarantee
when it times out on too-large tests.
Probabilistic Concurrency Testing %(PCT)
\cite{randomized-scheduler}
proposes to use random exploration of the state space and quantify the probability
that a bug may remain in some untested interleaving after a time-out,
although it is opaque to the user about which parts of her code were actually tested.
Meanwhile, ICB proposes to inform the user
of the maximum number of preemptions that were used to test any individual interleaving,
under the assurance that most bugs are likely to be found with fewer preemptions (\Cref{tab:icb-bounds}).
These guarantees could each be useful to developers in different scenarios,
and future work could combine the two approaches to provide both at once.
Iterative Deepening offers a clear benefit to the expert user
via the {\tt within\_function} command,
which enables her to restrict a test's scope to only the modules of a codebase she wishes to test.

Likewise, when full verification is not computationally feasible,
some jobs with data-race preemption points will inevitably time out,
and Quicksand cannot guarantee those races are
false positives or
benign, even though no bug was found.
In the ``Untested DR PPs'' column of \Cref{tab:drstatistix},
I show how many such candidates Quicksand (with Pure Happens-Before) could not verify in each test,
ultimately totaling 38\% of all data races in tests which timed out.
In prior work,
Portend introduced the {\em k-witness harmless} metric \cite{portend}
for heuristically classifying the likelihood that each data race lead to a failure or be benign.
Quicksand could incorporate this metric to guide the user's attention
to the unverified data races most likely to be worth her time.

In \cref{sec:quicksand-eval-partial} I presented partial verification results
measured in tens or hundreds of subset state spaces completed on average per test case.
However,
%On the other hand,
attempting to maximize the raw number of completed state spaces
is not necessarily the most user-friendly way to present partial verifications.
For starters, those numbers included small state spaces which were subsets of other state spaces also completed;
the user need not examine both subset and superset separately to understand what was tested.
Future work should at least perform basic set comparisons
to present only the non-redundant state spaces completed when time runs out.
For a further research challenge,
user studies could help to determine the most effective interface for presenting these partial results
from a software development perspective,
which I discuss further in \cref{sec:future-friendly}.

@@

quicksand is gr9

>>>>

Note that Iterative Deepening is a wrapper algorithm around stateless model checking.
A model checker is still used to test each state space, and other reduction techniques such as DPOR
are still applicable in each.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new preemption points reactively based on any runtime analysis.
Here
will focus on run-time data race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
The next section (\cref{sec:quicksand-soundness})
will prove that in fact,
in addition to statically-known synchronization preemption points,
this suffices to provide at least as strong verification guarantees as any other possible preemption point set.

@@

However, in many tests of even modestly-sized programs,
full verification is not feasible,
and focusing on the maximal state space alone is likely to be fruitless.
%
Hence, Iterative Deepening also allows for prioritizing subset jobs
based on number of preemption points, ETA, and whether data race candidates are included among their preemption points.
It relies on state-space estimation \cite{estimation}
to predict which jobs are likely to complete within a reasonable time,
before actually testing a large fraction of interleavings for each.
The overall goal is to decide automatically when to defer testing a state space,
so an inexpert user can provide only her total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.
Quicksand seeks to maximize completed state spaces,
as each one serves as a guarantee that all possible interleavings therein were tested;
\cref{sec:quicksand-discussion-partial} discusses some limitations of this approach.
The next three subsections will show how to schedule these smaller jobs
based on their preemption points and ETAs.

@@

For kernel-level testing, interrupt-disabling is analogous to locking,
so preemptions must also be introduced
just before a disable-interrupt opcode (on x86, {\tt cli})
and just after interrupts are re-enabled (on x86, {\tt sti}).
During data race analysis, {\tt cli} and {\tt sti} are treated as a single global lock
(note that {\tt cli}'d memory accesses can still race with others that have interrupts on).%
\footnote{Some kernels disable preemption without disabling interrupts,
which can be communicated to the model checker using manual annotations,
and must be treated similarly.
This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}
Quicksand is configured to begin with
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
$\{yield,cli\}$,
$\{yield,sti\}$,
and $\{yield,lock,$ $unlock,cli,sti\}$.
As a heuristic, it doesn't test every intermediate subset such as $\{lock,sti\}$,
which would result in $2^p$ jobs right off the bat,
although this could potentially be improved in future work (\cref{sec:quicksand-discussion-subsets}).

@@

\begin{figure}[t]
	\begin{center}
        \includegraphics[width=0.8\textwidth]{dr-jobs-v2.pdf}
	\end{center}
	\caption[Quicksand incorporates data races as new preemption points at run-time.]
		{Quicksand incorporates data races as new preemption points at run-time,
		by managing the exploration of multiple state spaces,
		communicating with each Landslide instance to receive ETAs, data races, and bug reports.
                %When an access pair is reported as a data race candidate,
                When a data race is reported,
                a new preemption point is added for each involved access,
                and new jobs are added for later testing,
                corresponding to different combinations of those with the existing preemption points.}
        \label{fig:new-dr-jobs}
\end{figure}

The new state spaces may expose a failure, in which case Iterative Deepening must stop and report a data-race bug,
or complete successfully, indicating a {\em benign} (i.e., false-positive) data race.
They may also uncover a new data-race candidate entirely in some alternate interleaving,
in which case we may iteratively advance to a superset state space which will preempt at both racing access pairs.
Being constrained by a CPU budget,
Iterative Deepening may time out before completing a data race's associated state space,
in which case the data race remains neither confirmed nor refuted.
%Depending on how much burden the implementation wants to impose on the user,
%it may then report it as a
%report a potential false positive that the user must handle
In such cases, Quicksand elects to impose some burden on the user
by reporting it as a potential false positive
and recommend that she investigate it by hand to judge for herself whether it be a bug.
\cref{sec:quicksand-discussion-partial} will discuss future opportunities for improving
debugging output in cases of such {\em partial verification}.
However, experience shows that this interactivity pays off:
in the next chapter's educational user study (\cref{sec:education-eval}),
one student reported during the survey that they used this recommendation,
combined with their own intuition,
to find a bug that Quicksand was not able to find alone (\cref{sec:education-reasons-worthwhile}).

@@

Iterative Deepening also accounts for the inherent inaccuracy of ETA estimates.
Line~1 heuristically scales up the time remaining to avoid suspending jobs too aggressively
in case their ETAs are actually overestimated.
Lines~12-15 account for the
%bizarre
possibility that among two suspended jobs,
%given two jobs,
%%$j_1,j_2$,
\GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$}
but
\GetETA{$j_1$} $>$ \GetETA{$j_2$}.
This may seem surprising,
but can often arise because estimates tend to get more accurate over time,
and $j_1$ perhaps ran much longer, on account of being overall smaller,
before becoming suspended.
In such scenarios,
the algorithm heuristically assumes the smaller job's ETA is more accurate,
in order to avoid repeatedly resuming larger jobs briefly only to find that their ETAs keep getting worse and worse.%
\footnote{In other words, it lets us avoid thrashing in Quicksand.} % ¯\_(ツ)_/¯

@@

As long as the allocation heap is properly synchronized,
a Pure Happens-Before analysis should identify a happens-before edge
between line~2's {\tt free()} and line~4's {\tt malloc()},
and report no race.
However, the upcoming evaluation will show that Limited Happens-Before retains some advantages over Pure
(\cref{sec:quicksand-eval-bugs}),
so it is useful to
%be able to
automatically suppress data-race candidates
that are certain to end up being false positives when reordered.
Such collisions could instead be avoided with a hacked allocator which never recycles memory,
but that could unacceptably impact performance in {\tt malloc()}-heavy tests.

The ability to disregard reallocation false positives is unique to Iterative Deepening.
When limited to a single test execution, suppressing any data race candidate matching this pattern is unsound.
Consider the more unusual program in Figure~\ref{fig:recycle-bug},
in which the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
Here, reordering lines~6-7 before line~3 will allow {\tt x} and {\tt x2} to race.
Discarding the data race report as a false positive after checking just this one execution
would overlook such a bug,
but Iterative Deepening is guaranteed to explore the alternate interleaving,
in which the true data race will show up without {\tt free()} and {\tt malloc()} interposing,
so it is safe to suppress at first, as I will prove in \cref{sec:quicksand-realloc}.
Moreover,
in the context of Iterative Deepening, being able to discard certain data race candidates
allows Quicksand to skip exploring some entire state spaces,
and hence run fewer Landslides overall;
this is analogous to DPOR's ability to skip equivalent interleavings within a single Landslide instance.
Upcoming in the evaluation, \cref{sec:quicksand-eval-verif}'s \Cref{tab:drstatistix}
will show how many redundant state spaces Quicksand is able to prune with this technique.

% TODO smae as above
\begin{figure}[t]
	\begin{center}
	\begin{tabular}{rll}
		& {\bf Thread 1} & {\bf Thread 2} \\
		1 & \texttt{publish(x);} & \\
		2 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
		3 & \texttt{\hilight{olivegreen}{free}(x);} \\
		4 & & \texttt{x2 = get\_published\_x();} \\
		5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
		6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
		7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
	\end{tabular}
	\end{center}
	% TODO: shortver
	\caption{If a single-pass Limited HB analysis discarded candidates matching the malloc-recycle pattern,
it would miss the bug in this adversarial program.}
	\label{fig:recycle-bug}
\end{figure}

@@

{\bf Assumptions.}
The proofs are built on a DPOR definition which assumes sequentially-consistent memory hardware.
All algorithms involved are assumed to operate on a machine model of a single globally-consistent execution trace,
which fundamentally cannot account for memory reordering nondeterminism.
%\cref{sec:quicksand-discussion} discusses this limitation further; % that's a lie..
For existing work on combining DPOR with relaxed memory, I refer the reader to \cite{tsopso}.
They also assume the Limited Happens-Before definition
for the data-race analysis.
I leave the case for Pure Happens-Before to future work,
although if I may appeal to intuition,
it requires only to show that for any data-race candidate
Limited Happens-Before reports in a given execution,
that Pure Happens-Before does not,
either it will be a false positive,
or the latter will find it in an alternate execution within the same state space,
or the latter will find a different data race that ultimately leads to a bigger state space in which the first one may be found,
much like a generalization of \cref{sec:quicksand-realloc}.

@@

\begin{itemize}
	\item {\bf Single state space, basic DPOR} ({\tt -C}):
		Runs a single instance of Landslide configured to preempt
		on all statically-known synchronization preemption points.
		Corresponds to dBug's approach \cite{dbug-ssv}.
		Never adds any new preemption points based on data race reports.
	\item {\bf Single state space, ICB} ({\tt -I}):
		Runs a single instance of Landslide with preemption points as above,
		but running Iterative Context Bounding with BPOR (\cref{sec:landslide-icb})
		instead of plain DPOR.
		Corresponds to CHESS's approach \cite{chess}.
		Requires either {\tt -C} or {\tt -M} (see below).
	\item {\bf Single state space, preempt-everywhere} ({\tt -0}):
		Runs a single instance of Landslide as above,
		but preempting on every shared memory access, not just synchronization.
		Corresponds to the approach of SPIN \cite{spin} and Inspect \cite{inspect};
		CHESS supports this mode as well with optional compiler instrumentation.
		Requires {\tt -C}; may optionally be combined with {\tt -I}.
	\item {\bf Maximal state space mode} ({\tt -M}):
		Runs the na\"{i}ve version of Iterative Deepening shown in \Cref{alg:algorithm0},
		i.e.,
		immediately abandons any state space whenever a superset of it exists.
		This results in always testing the maximal state space only, with no inherent parallelism,
		and optimizes for the fastest verification when the user has reason to believe no bugs will exist.
		No prior work implements this approach.
		Note that this mode was implemented after \cite{quicksand}'s publication,
		and I will feature it in the evaluation of transactional memory
		(\cref{sec:tm-eval}) rather than in this chapter.
\end{itemize}

@@

%\begin{figure}[p]
%	\begin{tabular}{c}
%        \includegraphics[width=0.71\textwidth]{../proposal/dowefindbugsfaster-v2.pdf} \\
%		\begin{tabular}{p{\textwidth}}
%                (a) Bugs found by Quicksand versus control experiments, measured in CPU time.
%                %as a function of elapsed CPU time.
%                Overall,
%                a more resource-fair comparison than (b),
%                although Quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\
%		\end{tabular}
%                \\
%                \\
%        \includegraphics[width=0.71\textwidth]{../proposal/dowefindbugsfaster-wallclock-v2.pdf} \\
%                %(b) Bugs found by elapsed wall-clock time.
%		\begin{tabular}{p{\textwidth}}
%                (b) Bugs found by Quicksand versus control experiments, measured in wall-clock time.
%                Quicksand is parallelized tenfold; the vertical line indicates its 1 hour limit. \\
%		\end{tabular}
%        \end{tabular}
%	\caption[Bug-finding performance comparison
%	between Quicksand and prior work.]
%	{Bug-finding performance comparison between
%        by several configurations of Quicksand and the single-state-space approaches.
%	Quicksand finds 125\% as many bugs
%        with data-race preemption points
%	at the 10-hour mark, compared to the best prior work approach.}
%        \label{fig:dowefindbugsfaster}
%\end{figure}

@@

Beyond finding new bugs and completing full verifications with data-race preemption points,
I evaluated Quicksand's performance for classifying data-race candidates in two ways:
its ability to check nondeterministic data races not reachable under a single-pass analysis (\cref{sec:quicksand-pps})
and its ability to suppress reallocation false positives (\cref{sec:quicksand-id-realloc}).
\Cref{tab:drstatistix} presents the results for this section.

@@

\Cref{fig:dr-falsenegs}
%compares the types of data-race candidates necessary to expose each data-race bug in the test suite.
visualizes the difference between single-pass and model-checking-enabled data race analysis.
The first and third series represent the bugs found using preemption points from single-pass data-race candidates only,
% not entirely true, as portend could be given data-race traces from an MC
%but they don't do it in their paper, so i feel comfortable making this claim
i.e., the state-of-the-art approach used by RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}.
The second and fourth series show all data-race bugs Quicksand found,
which includes the former type as well as new bugs involving nondeterministic races.
QS-Limited-HB found a nice 69 data-race bugs in total, % nice
15 of which %could not be found with single-pass data-race candidates alone.
required nondeterministic data-race preemption points to expose.
QS-Pure-HB is even more dependent thereupon, %on nondeterministic data-race preemption points,
%requiring nondeterministic data-race preemption points
requiring them in 21 cases among its 44 total data-race bugs.
Moreover, although the frequency
of these nondeterministic races varies across the different test cases
(for example, almost all in {\tt broadcast\_test} were nondeterministic; almost none in {\tt mutex\_test}),
they are still at least present in all tests,
meaning it is not just an issue of writing ``better'' test cases to avoid them.

@@

\subsection{Avoiding redundant work}

@@

\subsection{Preemption point subsets}
\label{sec:quicksand-discussion-subsets}

@@

In \cref{sec:quicksand-eval-bugs}, I evaluated the state-of-the-art approach's ability to find data-race-induced failures
by configuring a static predicate to preempt on any non-stack memory access
({\tt -0}, see \cref{sec:quicksand-impl-modes}).
This introduced hundreds of new preemption points on each new test execution,
with a prohibitive performance impact.
While this performance could be improved by
relaxing the preemption strategy,
instead using a static or single-pass analysis to find data-race candidates in advance \cite{portend},
that would sacrifice soundness of the verification guarantee, as I showed in \cref{sec:quicksand-eval-nondets}.
However, Quicksand itself could employ static data-race analysis such as RacerX \cite{racerx},
or single-pass dynamic analysis such as ThreadSanitizer \cite{tsan} in future work.
% These tools tend to err on the side of false positives,
% but also include .....
Any data-race candidates identified in advance could heuristically be included in Quicksand's initial seed preemption point sets
(\cref{sec:quicksand-initial-pps}),
enabling it to focus on the most suspicious races immediately,
rather than waiting for them to be identified after potentially many iterations of model checking.

\subsection{Partial verification}
\label{sec:quicksand-discussion-partial}

% Likewise,
When full verification is not computationally feasible,
some jobs with data-race preemption points will inevitably time out,
and Quicksand cannot guarantee those races are
false positives or
benign, even though no bug was found.
In the ``Untested DR PPs'' column of \Cref{tab:drstatistix},
I show how many such candidates Quicksand (with Pure Happens-Before) could not verify in each test,
ultimately totaling 38\% of all data races in tests which timed out.
In prior work,
Portend introduced the {\em k-witness harmless} metric \cite{portend}
for heuristically classifying the likelihood that each data race lead to a failure or be benign.
Quicksand could incorporate this metric to guide the user's attention
to the unverified data races most likely to be worth her time.
%
In \cref{sec:quicksand-eval-partial} I presented partial verification results
measured in tens or hundreds of subset state spaces completed on average per test case.
However,
%On the other hand,
attempting to maximize the raw number of completed state spaces
is not necessarily the most user-friendly way to present partial verifications.
For starters, those numbers included small state spaces which were subsets of other state spaces also completed;
the user need not examine both subset and superset separately to understand what was tested.
Future work should at least perform basic set comparisons
to present only the non-redundant state spaces completed when time runs out.
For a further research challenge,
user studies could help to determine the most effective interface for presenting these partial results
from a software development perspective,
which I discuss further in \cref{sec:future-friendly}.

Quicksand is not the first concurrency tester to provide a partial verification guarantee
when it times out on too-large tests.
Probabilistic Concurrency Testing (PCT)
\cite{randomized-scheduler}
proposes to use random exploration of the state space and quantify the probability
that a bug may remain in some untested interleaving after a time-out,
eschewing DPOR's depth-first search model to
instead sample broad cross-sections of large state spaces.
However, it proposes no alternate reduction strategy, making full verification impractical,
and furthermore is opaque to the user about which parts of her code were actually tested.
Meanwhile, ICB proposes to inform the user
of the maximum number of preemptions used to test any individual interleaving,
under the assurance that most bugs are likely to be found with fewer preemptions (\Cref{tab:icb-bounds}).
Iterative Deepening
%offers a clear benefit to the expert user
%via the {\tt within\_function} command,
%which enables her
allows the expert user
to restrict a test's scope via the {\tt within\_function} command to only the modules of a codebase she wishes to test.
These guarantees could each be useful to developers in different scenarios,
and future work could combine the three approaches to provide all benefits at once,
for example, using ETAs to heuristically decide when to switch between DPOR, ICB, and/or PCT in large state spaces,
as discussed further in \cref{sec:warpzone-heuristics}.

@@

quicksand is gr8
